---
title: "Using Optimal Transport theory to predict Occupational Mobility"
format:
  revealjs:
    smaller: true
    slide-number: true
    scrollable: false
    theme: solarized
    margin: 0
slide-number: true
incremental: true
widescreen: true
self-contained: true
execute:
  warning: false
  message: false
  echo: false
  error: false
---

```{r, include=FALSE}
library(tidyverse)
library(here)
skills <- read_rds("skills.rds")
results <- read_rds("results.rds")
```

## Introduction

* We know very little about occupational mobility.
* Anecdotally we know that people change careers, and using repeated cross-sectional Census NOC proportions we can observe aggregate shifts.
* Consider the 2011 Census age group 15-24, which 10 years later is the 2021 Census 25-34 age group.
* If we ignore the fact that the Census is a 25% sample (regarding occupation) and turn a blind eye to migration, then differences in NOC proportions must be due to inter-occupational mobility. 

---

## Changes in Occupation Proportions

```{r, fig.retina=2}
print(results$source_destination_plot[[1]])
```

---


## Changes in Occupation Proportions

```{r, fig.retina=2}
print(results$source_destination_plot[[2]])
```

---


## Changes in Occupation Proportions

```{r, fig.retina=2}
print(results$source_destination_plot[[3]])
```

---

## Changes in Occupation Proportions

```{r, fig.retina=2}
print(results$source_destination_plot[[4]])
```

---

## Changes in Occupation Proportions

* Greatest shifts in the occupation distribution occur between the age brackets 15-24 and 25-34: for many individuals this is when the make the transition from "job" to "career", likely after attaining some post secondary education.
* Both the *number* of occupations with a change in proportion greater than .005, as well as the *scale* of the changes in proportion diminish for older age brackets.

---

## Cost Minimizing Occupational Mobility

* To reiterate, we can not directly observe career changes, but we can infer they must occur based on shifts in occupational proportions between the two census. 
* There is an infinite set of occupational changes that would be consistent with both the initial and subsequent proportions. 
* Below we make use of the linear programming solution to the optimal transport problem to solve for the occupational transitions which minimizes "cost".

---

## Example

- Suppose goal is to end up with 4 evenly sized piles of dirt.
- Initially, the piles are arranged in a unit square, and have masses 1kg, 2kg, 3kg, 4kg.
- Cost is defined to be $\sum mass \times distance$.
- What is least costly way to end up with 4 piles of 2.5kg?

::: {.fragment}
```{r}
#| out.width: 50%
#| out.height: 50%
knitr::include_graphics(here("images","no_arrows.svg"))
```
:::

---

## Solution {.columns}

::: {.column width="50%"}

```{r}
#| out.width: 100%
#| out.height: 100%
knitr::include_graphics(here("images","just_dirt.svg"))
```

:::

::: {.column width="50%"}

* Move .5kg 1 unit of distance.
* Move 1kg $\sqrt2$ units of distance (hypotenuse)
* Move .5kg 1 unit of distance.
* $\min(cost)=1+\sqrt2$. 

:::

---

## How is this related to labour market?

* Rather than piles of dirt, consider 512 "piles" of workers.
* Specifically, consider the NOC proportions of those 15-24 in 2011: this corresponds to the size of the original "piles".
* Likewise, the NOC proportions of those 25-34 in 2021 are the desired "piles".
* There is an infinite number of ways to move workers to attain the desired piles.
* The optimal transport problem picks the least costly solution from the set of possible solutions.
* The solution provides the lower bound of (costly) occupational transitions.

---

## Distance between occupations

* We use 161 occupational characteristics from ONET to measure the distance between occupations.
* Principal component analysis is used to reduce the dimensionality of the data. 
* For our distance metric we calculate the Euclidean distance on the first 10 principal components.

---

## Screeplot


```{r, fig.retina=2}
eig <- factoextra::get_eig(skills[["onet_pca"]]) |>
head(n=10)
eig$Dim <- seq_len(nrow(eig))

ggplot(eig, aes(x = factor(Dim))) +
geom_bar(aes(y = variance.percent), stat = "identity") +
geom_line(aes(y = cumulative.variance.percent), size = 0.8, group = 1) +
geom_point(aes(y = cumulative.variance.percent)) +
scale_y_continuous(
name = "Individual % variance",
sec.axis = sec_axis(~ ., name = "Cumulative % variance")
) +
labs(title="First 10 components capture over 80% of the variability", 
     x = "Principal Component") +
theme_minimal()
```

---

## Distances Between Occupations {.columns}

::: {.column width="50%"}

```{r, fig.height=5, fig.width=5, cache=TRUE}
heatmaply::heatmaply(
  skills[["skills_noc_dist"]],
  show_dendrogram = c(FALSE, FALSE),
  showticklabels = c(FALSE, FALSE)
)|>
  plotly::layout(autosize = TRUE) |>
  plotly::config(responsive = TRUE)
```

:::

::: {.column width="50%"}

* Along the diagonal the distance is zero between an occupation and itself.
* Clusters of occupations with similar skill profiles are located in darker rectangles along the diagonal.

:::

---

## Transitions: 15 to 24 years -> 25 to 34 years

<video width="800" controls>
  <source src="15_to_24_years -> 25_to_34_years.mp4" type="video/mp4">
  Your browser does not support the video tag.
</video>

---

## Transitions: 25 to 34 years -> 35 to 44 years

<video width="800" controls>
  <source src="25_to_34_years -> 35_to_44_years.mp4" type="video/mp4">
  Your browser does not support the video tag.
</video>

---

## Transitions: 35 to 44 years -> 45 to 54 years

<video width="800" controls>
  <source src="35_to_44_years -> 45_to_54_years.mp4" type="video/mp4">
  Your browser does not support the video tag.
</video>

---

## Transitions: 45 to 54 years -> 55 to 64 years

<video width="800" controls>
  <source src="45_to_54_years -> 55_to_64_years.mp4" type="video/mp4">
  Your browser does not support the video tag.
</video>

---

## Summary of the flows

* Maybe animations are not your thing.
* Below we aggregate all the NOC flows to the TEER level, allowing us to visualize using (static) alluvial plots.

--- 

## TEER Flows

```{r, fig.retina=2}
results$teer_alluvium_plot[[1]]
```

---


## TEER Flows

```{r, fig.retina=2}
results$teer_alluvium_plot[[2]]
```

---


## TEER Flows

```{r, fig.retina=2}
results$teer_alluvium_plot[[3]]
```

---


## TEER Flows

```{r, fig.retina=2}
results$teer_alluvium_plot[[4]]
```

---





